Creating optimal conditions for reproducible data analysis in R with `fertile`

### 2020-02-22 fertile.Rmd
reformatting package names

There appear to be no existing tools that cater to analysts looking for an easy-to-learn, easy-to-implement, fast way to obtain a broad overview of their projects' reproducibility.
To address this, we present an R package called [`fertile`](https://github.com/baumer-lab/fertile) [^1]:{The authors gratefully acknowledge contributions from Hadley Wickham, Jenny Bryan, Greg Wilson, Edgar Ruiz, and other members of the `tidyverse` team.}, a low barrier-to-entry package which focuses on a series of common mistakes programmers make while conducting data science research in R.

@@

Much of the available literature is focused on the methods for achieving reproducibility in specific disciplines, though much of this information is generalizable to all areas of scientific research. There is also a small body of work focusing on reproducibility within the R community specifically. Some of this work is in the form of academic papers, while other literature comes from the blog posts and websites of several highly-regarded R-focused websites and researchers.  In this section, we will discuss the reproducibility recommendations from both the formal and informal publications, which together influenced which features were included in `fertile`.

There is also a small existing body of R packages focused on providing tools to assist researchers with achieving reproducibility. We will consider the functionality of these packages, identifying `fertile`'s unique role and functionality.

@@

@@

As we have seen, many existing R packages focusing on reproducibility are designed in a way such that they can only be used prior to the start of a project, often for the purposes of creating an organized file structure or ordered file hierarchy. 
`fertile`, on the other hand, can be used at any stage in the data analysis workflow. It is designed to be used on data analyses organized as R Projects (i.e. folders containing a `.Rproj` file). Once an R Project is created, `fertile` can provide benefits at any time, both while coding as well as after the fact.
`fertile` operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems). 
`fertile` is available for download at [https://github.com/baumer-lab/fertile](https://github.com/baumer-lab/fertile)

@@

Proactively, the package identifies potential mistakes as they are made by the user and outputs an informative message as well as a recommended solution. For example, `fertile` catches when a user passes a potentially problematic file path---such as an absolute path, or a path that points to a location outside of the project directory---to a variety of common input/output functions.

@@

```{r, include = FALSE, message = FALSE}
readr::write_csv(mtcars, "~/Desktop/my_data.csv")
library(fertile)
library(tidyverse)
```

```{r, message = FALSE}
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
```

@@

`fertile` is even more aggressive with functions (like `setwd()`) that are almost certain to break reproducibility---it causes them to throw errors.

@@

The proactive features are activated immediately after loading the `fertile` package and require no additional effort by the user.

In addition to the interactive warning system, `fertile` provides several useful utility functions. Among other things, these include functions to check the type of a file and a way to create a copy of a project in a temporary directory.

@@

Retroactively, `fertile` analyzes potential obstacles to reproducibility in an RStudio Project (i.e., a directory that contains an `.Rproj` file), including the directory structure, the analyst's use of file paths, randomness, etc. `fertile}` creates reproducibility reports that identify potential mistakes and provide recommendations for remedies. For example, `fertile` might identify the use of randomness in code and recommend setting a seed.

Users can access the majority of `fertile`'s retroactive features through two primary functions.

The `proj_check()` function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that to not have a call to `set.seed()` present. Subsets of the fifteen tests can be invoked using the select helper functions from `dplyr` and the `proj_check_some()` function.

@@

`fertile` is designed in such a way as to be useful to users of all backgrounds and experiences. In this section, we provide two sample use cases demonstrating situations where users could benefit significantly from using the package.

@@

Susan doesn't understand why it didn't work on the professor's computer when it did on her own. The professor recommends that she install `fertile` and run `proj_check()` on her assignment. She does this and gets a message informing her she used an absolute path to open her dataset, and she should use a relative path. She looks up what this means and then uses the new information to update her assignment. She resubmits and her second attempt is successful.

@@

Unsure of the best way to achieve this structure, Emma goes online to find help. In her search, she comes across `fertile`. She downloads the package, and in only a handful of commands, she identifies and removes excess files in her directory and automatically organizes her files into a structure reminiscent of an R package. She now meets the guidelines for the journal and can submit her research.

@@

In an effort to understand the package's effectiveness, we also share preliminary results from a randomized, controlled experiment conducted on an undergraduate introductory data science course [^2]:{This study is approved by Smith College IRB, Protocol #19-032}. The purpose of the study is to determine whether `fertile` helps students produce data science research that is more likely to be reproducible.

@@

`fertile` is an R package that lowers barriers to reproducible data analysis projects in R. 
The features of `fertile` can be accessed almost effortlessly, making it easy for data analysts of all skill levels and backgrounds to gain a better understanding of how to make their work reproducible.
`fertile` is designed to educate the user about why the mistakes are problematic and how to fix them, promoting a greater understanding of reproducibility concepts in its users.
It is written for a wide audience, simple enough to be used by students in an introductory data science course, but still helpful to experienced analysts.
`fertile` also addresses a human challenge of reproducibility. In the moment, it often feels easiest to take a shortcut---to use an absolute path or change a working directory. However, when considering the long term path of a project, spending the extra time to improve reproducibility is worthwhile. `fertile`'s user-friendly features can help data analysts avoid these harmful shortcuts with minimal effort.

>>>>

There appear to be no existing tools that cater to analysts looking for an easy-to-learn, easy-to-implement, fast way to obtain a broad overview of their projects' reproducibility.
To address this, we present an R package called [**fertile**](https://github.com/baumer-lab/fertile) [^1]:{The authors gratefully acknowledge contributions from Hadley Wickham, Jenny Bryan, Greg Wilson, Edgar Ruiz, and other members of the `tidyverse` team.}, a low barrier-to-entry package which focuses on a series of common mistakes programmers make while conducting data science research in R.

@@

Much of the available literature is focused on the methods for achieving reproducibility in specific disciplines, though much of this information is generalizable to all areas of scientific research. There is also a small body of work focusing on reproducibility within the R community specifically. Some of this work is in the form of academic papers, while other literature comes from the blog posts and websites of several highly-regarded R-focused websites and researchers.  In this section, we will discuss the reproducibility recommendations from both the formal and informal publications, which together influenced which features were included in **fertile**.

There is also a small existing body of R packages focused on providing tools to assist researchers with achieving reproducibility. We will consider the functionality of these packages, identifying **fertile**'s unique role and functionality.

@@

Much of the available literature focuses on file structure, organization, and naming, and **fertile`**s features are consistent with this. The *ideal* file structure is not agreed upon by academics, though there are some recommendations focused on R that are available. *Packaging data analytical work reproducibility using {R} (and friends)* by Marwick, Boettiger, and Mullen provides the framework for file structure that **fertile** is based on: a structure similar to that of an R package, with an `R` folder, as well as `data`, `data-raw`, `inst` and `vignettes`.

@@

As we have seen, many existing R packages focusing on reproducibility are designed in a way such that they can only be used prior to the start of a project, often for the purposes of creating an organized file structure or ordered file hierarchy. 
**fertile**, on the other hand, can be used at any stage in the data analysis workflow. It is designed to be used on data analyses organized as R Projects (i.e. folders containing a `.Rproj` file). Once an R Project is created, **fertile** can provide benefits at any time, both while coding as well as after the fact.
**fertile** operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems). 
`fertile` is available for download at [https://github.com/baumer-lab/fertile](https://github.com/baumer-lab/fertile)

@@

Proactively, the package identifies potential mistakes as they are made by the user and outputs an informative message as well as a recommended solution. For example, **fertile** catches when a user passes a potentially problematic file path---such as an absolute path, or a path that points to a location outside of the project directory---to a variety of common input/output functions.

@@

```{r, include = FALSE, message = FALSE}
readr::write_csv(mtcars, "~/Desktop/my_data.csv")
library(tidyverse)
library(fertile)
```

```{r, message = FALSE}
file.exists("~/Desktop/my_data.csv")
read_csv("~/Desktop/my_data.csv")
```

@@

**fertile** is even more aggressive with functions (like `setwd()`) that are almost certain to break reproducibility---it causes them to throw errors.

@@

The proactive features are activated immediately after loading the **fertile** package and require no additional effort by the user.

In addition to the interactive warning system, **fertile** provides several useful utility functions. Among other things, these include functions to check the type of a file and a way to create a copy of a project in a temporary directory.

@@

Retroactively, **fertile** analyzes potential obstacles to reproducibility in an RStudio Project (i.e., a directory that contains an `.Rproj` file), including the directory structure, the analyst's use of file paths, randomness, etc. **fertile** creates reproducibility reports that identify potential mistakes and provide recommendations for remedies. For example, **fertile** might identify the use of randomness in code and recommend setting a seed.

Users can access the majority of **fertile**'s retroactive features through two primary functions.

The `proj_check()` function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that to not have a call to `set.seed()` present. Subsets of the fifteen tests can be invoked using the select helper functions from **dplyr** and the `proj_check_some()` function.

@@

**fertile** is designed in such a way as to be useful to users of all backgrounds and experiences. In this section, we provide two sample use cases demonstrating situations where users could benefit significantly from using the package.

@@

Susan doesn't understand why it didn't work on the professor's computer when it did on her own. The professor recommends that she install **fertile** and run `proj_check()` on her assignment. She does this and gets a message informing her she used an absolute path to open her dataset, and she should use a relative path. She looks up what this means and then uses the new information to update her assignment. She resubmits and her second attempt is successful.

@@

Unsure of the best way to achieve this structure, Emma goes online to find help. In her search, she comes across **fertile**. She downloads the package, and in only a handful of commands, she identifies and removes excess files in her directory and automatically organizes her files into a structure reminiscent of an R package. She now meets the guidelines for the journal and can submit her research.

@@

In an effort to understand the package's effectiveness, we also share preliminary results from a randomized, controlled experiment conducted on an undergraduate introductory data science course [^2]:{This study is approved by Smith College IRB, Protocol #19-032}. The purpose of the study is to determine whether **fertile** helps students produce data science research that is more likely to be reproducible.

@@

**fertile** is an R package that lowers barriers to reproducible data analysis projects in R. 
The features of **fertile** can be accessed almost effortlessly, making it easy for data analysts of all skill levels and backgrounds to gain a better understanding of how to make their work reproducible.
**fertile** is designed to educate the user about why the mistakes are problematic and how to fix them, promoting a greater understanding of reproducibility concepts in its users.
It is written for a wide audience, simple enough to be used by students in an introductory data science course, but still helpful to experienced analysts.
**fertile** also addresses a human challenge of reproducibility. In the moment, it often feels easiest to take a shortcut---to use an absolute path or change a working directory. However, when considering the long term path of a project, spending the extra time to improve reproducibility is worthwhile. **fertile**'s user-friendly features can help data analysts avoid these harmful shortcuts with minimal effort.
